A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-09 graph_syntax.tex

\section{Requirements for New Graph Syntax}
The ultimate goal of the new graph--based syntax presented here is to be able to 
fully describe the general structure of an MDAO problem indpendently of any solution information, 
while still being able to accomodate the more specific case when a solution 
strategy is applied. In order to achieve that goal 
the graph needs to accomodate a number of constructs of MDAO problems:

\begin{itemize}
    \item Analysis tools and connections between them
    \item Design variables, objectives, and constraints
    \item Local and global properties
    \item Coupling between analyses
    \item Multi-fidelity analyses
\end{itemize}

Beyond those basic constructs, there are also three phases of a design process that 
all need to be representable with the new syntax. Firstly there is the initial problem definition
phase where the specific analysis tools and design goals are identified. At the end of this phase, 
a single formal problem formulation is selected specifying design variables, constraints, objectives, 
analysis tools, etc. Lastly some specifc procedure for solving the problem is selected, for example 
picking an MDAO optimization architecture. Using the proposed graph syntax, each of these phases 
can be represented with a related graph.

\begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem  (FPG)
    \item Problem Solution Graph (SPG)
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all 
analysis codes being considered and all possible connections between them also present. The second graph 
is \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and nodes to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPGs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPGs that could be reached 
untill only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
different solution strategies. If you consider the size of a graph to be the sum of all its
edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
In other words the FPG is obtained from the MCG by removing nodes and edges, 
and the PSG is obtained from the FPG by adding nodes and edges.

@@

\subsection{Node and Edge Types}

There are four node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[expression:] simple algebreic combinations of variables (such as for 
  constraint of objective functions)
\item[driver:] control logic capable of managing iteration
\end{description}

In addition to the four node types, there are two edge types:

@@

\subsection{Rules for Nodes and Edges}

The driver node and the implicit edge are only allowed to be pressent in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed connection edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

@@

\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower in-degree limit}
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}
These user-specified limits govern the number of fixed and connection edges that 
may be directed into a node for a valid graph. Consider a variable node $v$ with
$\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, that node must have exactly one
incomming edge or the graph is invalid. Two possiblities exist for voilating 
the limit conditions: 
\begin{description}
  \item[hole: ] The number of incomming edges is less than $\txt{deg}_l^-(v)$. 
    \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}

  \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
    \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \label{e:collision} \end{equation}
\end{description}

\section{Graph Representation of MDAO Constructs}

\subsection{Analysis Blocks and Connections}

Analysis tools take in a number input variables, then perform some work to calculate 
the values for their respective outputs. In an MDAO graph, this process is 
represented by an group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig. \ref{f:analysis block}, 
the analysis block contains two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to proper output values. If constructing a weighted graph, the weight calculation edge 
provides the opportunity to computational cost. However, it is also allowed to skip 
this edge and have both incomming and outgoing edges to variables through a single model 
node for a given analysis block. Although analysis blocks are comprised of multiple nodes and edges, since all 
the edges within them are fixed, they become fixed sub-graphs within a graph.

@@

Via the analysis block structure, we can also distinguish between input and ouptut
variable nodes. Inputs are any variable nodes that have an outgoing edge into a model 
node. Conversely, outputs are any variable nodes that have an incoming edge from a model node. 
Note that by this definition, inputs and outputs can only exist when variable nodes poses an 
edge joining them to a model node. In Fig. \ref{f:analysis block} these are labeled out as 
local inputs and local outputs. The variables are local because they are explicitly tied
to the analysis block and have no relevance without it.

@@

\subsection{Design Variables, Objectives, and Constraints}

Design variables, objectives, and constraints all require external decisions 
to be made about a given system to identify them. Consider a notional problem 
given by the graph in Fig. \ref{f:designvars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the 
output, $x$, of $A$ is connected to the input, $x$, of $B$ x can't actually be a 
design variable. The inputs $w$ and $y$ have no incomming edges and could each 
be selected as design variables.

\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

More formally, any input variable that can be defined as a hole by Eqn. \ref{e:hole}
would be considered a design variable in the graph. When working with a MCG or 
FPG, only connection edges are allowed, in which case a design variable 
is simply a variable node with no incomming edges. For a valid PSG 
any design variable node would have at least one, but possibly multiple, incomming 
implicit edges from one or more driver nodes.

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:designvars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer identifies variables 
or groups of variables and assigns a simple relationship between them.

>>>>

\section{Requirements for New Graph Syntax}
The ultimate goal of the new graph--based syntax presented here is to be able to 
fully describe the general structure of an MDAO problem indpendently of any solution information, 
while still being able to accomodate the more specific case when a solution 
strategy is applied. In order to achieve that goal 
the graph needs to accomodate a number of constructs of MDAO problems: 
\begin{itemize}
    \item Analysis tools and connections between them
    \item Design variables, objectives, and constraints
    \item Local and global properties
    \item Coupling between analyses
    \item Multi-fidelity analyses
\end{itemize}

Beyond those basic constructs, there are also three phases of a design process that 
all need to be representable with the new syntax. Firstly there is the initial problem definition
phase where the specific analysis tools and design goals are identified. At the end of this phase, 
a single formal problem formulation is selected specifying design variables, constraints, objectives, 
analysis tools, etc. Lastly some specifc procedure for solving the problem is selected, for example 
picking an MDAO optimization architecture. Using the proposed graph syntax, each of these phases 
can be represented with the following three graphs:
\begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem  (FPG)
    \item Problem Solution Graph (SPG)
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all 
analysis codes being considered and all possible connections between them also present. The second graph 
is the \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and nodes to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPGs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPGs that could be reached 
until only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
different solution strategies. If you consider the size of a graph to be the sum of all of its
edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
In other words the FPG is obtained from the MCG by removing nodes and edges, 
and the PSG is obtained from the FPG by adding nodes and edges.

@@

\subsection{Node and Edge Types}
We now define different types of nodes and edges with specific properties to compose graphs used to represent an MDAO problem. 
There are three node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[driver:] control logic capable of managing iteration
\end{description}

In addition to the three node types, there are two edge types:

@@

\subsection{Rules for Nodes and Edges}
There are specific rules for the usage of these nodes and edges.
The driver node and the implicit edge are only allowed to be present in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to these restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed connection edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

@@

\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower in-degree limit} as
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}
These user-specified limits govern the number of connection edges that may be directed into a variable
node for a valid graph. Consider a variable node $v$ with
$\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ must have exactly one
incomming explicit edge or the graph is invalid. Two possiblities exist for voilating 
these limit conditions: 
\begin{description}
  \item[hole: ] The number of incomming edges is less than the lower indegree limit:
    \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}

  \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
    \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \end{equation}
\end{description} 
These violations will be managed in order to seek a working problem formulation.

\section{Graph Representation of MDAO Constructs}
%As discussed in Sec.((intro)), the digraph must be able to describe the individual aspects of an MDAO problem.
\subsection{Analysis Blocks and Connections}
Analysis tools take in a number input variables and then perform some work to calculate 
the values for their respective outputs. In an MDAO graph, this process is 
represented by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig. \ref{f:analysis block}, 
the analysis block contains two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to proper output values. If constructing a weighted graph, the weight calculation edge 
provides the opportunity to computational cost. However, it is also allowed to skip 
this edge and have both incomming and outgoing edges to variables through a single model 
node for a given analysis block. Although analysis blocks are comprised of multiple nodes and edges, since all 
the edges within them are fixed, they become fixed sub-graphs within a graph.

@@

Via the analysis block structure, we can also distinguish between input and ouptut
variable nodes, as shown in Fig.~\ref{f:analysis block}. Inputs are any variable nodes that have an outgoing edge into a model 
node. Conversely, outputs are any variable nodes that have an incoming edge from a model node. 
Note that by this definition, inputs and outputs can only exist when variable nodes poses an 
edge joining them to a model node. These are labeled out as 
local inputs and local outputs. The variables are local because they are explicitely tied
to the analysis block and have no relevance without it.

@@

\subsection{Design Variables}
Design variables require external decisions 
to be made about a given system to identify them. Consider 
a notional problem given by the graph in Fig. \ref{f:designvars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the variable $x$ is connected from $A$ to $B$,
 $x$ can't actually be a design variable. On the other hand, the inputs $w$ and $y$
have no incomming edges and could each be selected as design variables. 
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

Given a MCG with the ${deg}_l^-(v)=1$ for all nodes, there will be a set of holes. 
Each hole represents a node that could potentially become a design variable. 
It is up to the designer to examine that variable and decide if it is appropriate to 
use it as a design variable. If a variable should be a design variable, then 
set it's ${deg}_l^-(v)=0$. However, some variable holes might not make good design 
variables. For instance, if your aircraft mission analysis code requires a drag input 
which ends up being a hole then it's likely that some new analysis tool is needed
to predict drag before the design process can continue. In that case, you leave
${deg}_l^-(v)=1$ and find some tool to fill the hole in graph.

So by definition, a design variable is any variable node with ${deg}_l^-(v)=0$, 
When working with a MCG or FPG, only connection edges are allowed, so a design 
variable in either graph will be a variable with no edges. For a valid PSG 
any design variable node would have at least one, but possibly multiple, incomming 
implicit edges from one or more driver nodes.

\subsection{Objectives, and Constraints}
Objectives and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
algebraic expression to form a objective function. For example, the problem from 
Fig. \ref{f:designvars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer compiles an
algebraic expression that is a function of existing variable nodes in the graph.
